Less Funding, More Lecturers, and Fewer Professors: \\ \vspace{0.1cm} Stagnating State Funding for Higher Education and its Effect on Faculty

### 2024-03-15 text/sections/01-introduction.tex
updated

It is a priori unclear how faculty may be affected, as there are various margins that universities may to respond to funding cuts.
On one hand, faculty may be negatively affected by universities cutting salaries or laying-off incumbent faculty.
This happened in 2010 when the University of Illinois System furloughed university staff following budget cuts after the Great Recession \citep{furlough2010}.
Similarly, faculty hiring committees are often cancelled or disrupted in response to funding cuts, meaning there are fewer faculty entering a department to replace those leaving.
On the other hand, lecturers could be hired to fill in teaching duties, on shorter and cheaper contracts with fewer employment protections.
But it is not certain that universities would hire more lecturers in the face of budget cuts.
Public university administrators, motivated by increased competition between unviersity for student enrolment \citep{hoxby2009changing,urquiola2020markets},may notice that their rankings are negatively affected by falling professor--lecturer ratios, and so limit departments from hiring more lecturers to replace leaving/retiring professors.
Administrators could even start courting private funding or increase tuition prices to fill funding gaps \citep{bound2019public}, possibly leaving faculty unaffected by state funding cuts.
These are some ways that the long-run stagnation in state funding would not guaratee a substitution towards lecturers at US public universities.

>>>>

It is a priori unclear how faculty may be affected, as there are various margins that universities may to respond to funding cuts.
On one hand, faculty may be negatively affected by universities cutting salaries or laying-off incumbent faculty.
This happened in 2010 when the University of Illinois System furloughed university staff following budget cuts after the Great Recession \citep{furlough2010}.
Similarly, faculty hiring committees are often cancelled or disrupted in response to funding cuts, meaning there are fewer faculty entering a department to replace those leaving.
On the other hand, lecturers could be hired to fill in teaching duties, on shorter and cheaper contracts with fewer employment protections.
But it is not certain that universities would hire more lecturers in the face of budget cuts.
Public university administrators, motivated by increased competition \citep{hoxby2009changing,urquiola2020markets}, may notice that their rankings are negatively affected by falling professor--lecturer ratios, and so limit departments from hiring more lecturers to replace leaving/retiring professors.
Administrators could even start courting private funding or increase tuition prices to fill funding gaps \citep{bound2019public}, possibly leaving faculty unaffected by state funding cuts.
These are some ways that the long-run stagnation in state funding would not guarantee a substitution towards lecturers at US public universities.
